---
title: "Numbers into English Words"
author: "Bill Venables"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    includes:
      in_header: header.tex
vignette: >
  %\VignetteIndexEntry{Numbers into English Words}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      comment = "")
library(english)
```

## Genesis and development

In answer to a question on R-help John Fox provided an elegant R
function to translate integers into English numbers.  The present
package extends this code to an S3 class, with constructor functions and
methods to make this original idea more conveniently available.

The function `as.english` is intended to provide a parallel
facility to the function `as.roman` in the `utils` package.

The main purpose of the package is to present an interesting programming
example rather than to solve a likely real problem, though there could
well be some applications in unusual contexts.

Note added in Version 1.1-4.  The two small helper functions `words`
and `Words` are included to facilitate inline code inserts in
R markdown files.  See the help files for examples.  The `ordinal`
function produces character strings and may be used directly in inline
code inserts.  Use 
`` `r '\x60r words(103)\x60'` ``
rather than
`` `r '\x60r english(103)\x60'` ``
in R markdown files.

Note added in Version 1.2-0.  The function `indefinite` added for
algorithmically including an apprporiate indefinite article in a
document insert.  Based on a suggestion of Anne Pier Salverda.

The capitalised version, `Indefinite`, was added in 1.2-1.

This vignette was added in 1.2-2.

## Examples

The main use envisaged for these tools is for code inserts in either
R markdown or R noweb documents.  It is easier here to demonstrate their
function in plain code chunks, as below.

The resolution of an integer into words depends on the style of English
chosen.  English (UK) style inserts the conjunction *and* in various places
where American (USA) style omits it.  Wheter or not this happens is
governed by the `english.UK` option, (a logical variable, `TRUE` for UK
English), but if this option is unset, a guess is made as to which style
is needed depending on the user's locale.

```{r, results = "hold"}
soldiers <- 10006
oldOpt <- options(english.UK = TRUE)
cat("The Duke of York had approximately", words(soldiers), "men.\n")
cat("How many did you say?  ", Words(soldiers), ", approximately.\n", sep = "")
```
```{r, results = "hold"}
options(english.UK = FALSE)
cat("The Duke of York had approximately", words(soldiers), "men.\n")
cat("How many did you say?  ", Words(soldiers), ", approximately.\n", sep = "")
options(oldOpt)
```

Ordinal numbers are handled as well.

```{r}
days <- 1:6
cat(paste("\nOn the", ordinal(days), "day of Christmas..."))
```

The indefinite article, "a" or "an", can be algorithmically chosen and prepended,
either in lower case or capitalised:

```{r}
steps <- 7:11
cat(paste0("\nThis is ", indefinite(steps), "-step process..."))
cat(paste0("\nThis is ", indefinite(steps, words = FALSE), "-step process..."))
cat(paste0("\n", Indefinite(ordinal(steps)), " step of the process is..."))
```

Roman notation is provided by the `utils::as.roman` function in a similar way:

```{r, results = "hold"}
numbers <- c(1:10, 1999, 2019)
punct <- c(rep(",", 10), " and",".")
cat(c("In Roman notation:", 
      paste0("\n\t", Words(numbers), " is written as \"",
             utils::as.roman(numbers),"\"", punct)))
cat("\nDoing arithmetic in Roman notation can be difficult.")
```

Some arithmetic operations are allowed for `english` objects, but to make
sense the result should be in whole numbers.

```{r}
english(100) + (-5):5
```


